Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-06-29 agujournaltemplate.tex
Ago Review Intro revised

    % Phytoplankton is important
    Phytoplankton provide the foundation for marine food webs and play a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}. This paraphyletic community consists of a diverse group of organisms  exhibiting highly variable eco-physiological properties \cite{appeltans_magnitude_2012}. The short generation time and high diversity allow phytoplankton communities to respond relatively quickly to environmental changes, making them suitable indicators of ecosystem state \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
    Coastal margins, where phytoplankton typically thrive, constitute only \qty{\sim 10}{\%} of the global ocean surface  but are responsible for an estimated \qty{40}{\%} of carbon sequestration to the seafloor \cite{yool_examination_2001, mullerkarger_importance_2005}.

    Data from remote sensing, suggest a global increase in the frequency and spatial extent of coastal phytoplankton blooms over the last 20 years, but regional effects are highly variable \cite{dai_coastal_2023} and difficult to capture.
    Climate change is predicted to have a lasting impact on phytoplankton communities, including long-term shifts in community composition and alterations of diversity \cite{acevedo-trejos_glimpse_2014, boyd_biological_2016, henson_future_2021}. 
    Long-term time series of biological observations are crucial for understanding how marine ecosystems respond to short-termcyclic processes and to long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}. Such time series are rare, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.

    A tropical coastal location that has been studied intensively in the last decades is the CARIACO basin, where a wide range of physical and biogeochemical parameters have been consistently collected from 1995 to 2017 \cite{muller-karger_scientific_2019}. 
    % Describe CARIACO location
    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within one of the largest anoxic basins in the world oceans \cite{edgcomb_accessing_2011}. The basin is a continental shelf depression comprising two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}. The water below the sill is poorly ventilated. This underwater topography in combination with a high surface productivity, associated to aerobic degradation of fast sinking organic matter, creates permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}. Due to these unique properties, the Cariaco Basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of past climate \cite{hughen1996nature} and serving as a testbed for understanding the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.

    % Describe general seasonal cycle with functional groups etc.
    The Cariaco Basin is characterised by a marked seasonal cycle of upwelling, which is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward. A smaller wind-driven secondary upwelling event  occurs often in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}. During the upwelling period, the supply of deeper nutrient-rich waters supports high biological productivity, and the phytoplankton community is dominated by diatoms \cite{romero_seasonal_2009}, with dinoflagellates, coccolithophorids, and nanoplankton also contributing. In the low-wind, rainy season, the Cariaco Basin exhibits oligotrophic conditions and the fraction of large phytoplankton cells is reduced, albeit nano- and picophytoplankton cells are present in relatively high numbers throughout the year \cite{lorenzoni_characterization_2015}.

    % PARAGRAPH THREE - CHANGES IN CARIACO
    Since the inception of the time series in 1995, the ecosystem of the Cariaco Basin has undergone marked changes. \citeA{taylor_ecosystem_2012} analyzed the time series data up to 2010 and found significant decadal trends of increasing sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors attributed to global warming. During the initial phase of the time series (1996–2004), the Cariaco Basin underwent consistent and pronounced seasonal upwelling cycles, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}. After 2004, a significant reduction in the strength of trade winds during the upwelling period \cite{taylor_ecosystem_2012} caused a regime change in the physical and biogeochemical system. This strongly affected the phytoplankton community and the whole ecosystem. Microscopy cell counts indicated a drastic reduction of large phytoplankton, particularly diatoms, as confirmed, albeit not with the same high significance, by analysis on pigment data  \cite{pinckney_phytoplankton_2015}. In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, groups commonly occurring in stratified water columns \cite{lorenzoni_characterization_2015}. An analysis of pigment data indicates a deepening of the chlorophyll maximum, an increase in diversity, and a reduction in seasonal variability of community composition \cite{pinckney_phytoplankton_2015}. During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, coinciding with an increase in zooplankton biomass \cite{taylor_ecosystem_2012}. Interestingly, changes in the phytoplankton community did not result in a significant change in the export of particulate organic carbon (POC) to depth, as measured by sediment traps deployed during the time series \cite{pinckney_phytoplankton_2015}.

    % PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
    The shift in phytoplankton community composition associated to increased temperature and to reduced upwelling intensities, does not appear to be a permanent change in regime. Data from 2012 to 2017, the end phase of the CARIACO time series, indicates a return to strong upwelling intensities and to high numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}. The return to conditions similar to those of the late 1990s coincides with a recovery in sardine catches observed in 2016 and 2017 in an area southeast of Isla de Margarita \cite{gomez_gaspar_variacion_2025}. 
    The impact of large-scale climatic processes on the observed shifts in the phytoplankton community remains unclear. The Atlantic Multidecadal Oscillation (AMO) has emerged as an important factor influencing the climate system in the region and the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}. The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
    The El Niño Southern Oscillation (ENSO) is teleconnected to the Caribbean Sea given that during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures increase and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. However, using the time-series available only until 2010, \citeA{taylor_ecosystem_2012} found no significant correlation between MEI v.2 and in-situ observations. In addition, the AMO index was included in the analyses of the Cariaco Basin.

    % What drives the changes??? Bottom-up/ Top-down ... CLIMATE 
    \citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as "unidirectional trends driven by anthropogenic climate change", and indicated that they may "simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation".
    Previous analyses of phytoplankton community diversity in the Cariaco Basin did not include time series data beyond 2011, and no detailed investigations have been undertaken yet on possible changes in diversity based on microscopy cell counts.
    Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period 1995\,--\,2017, 2) changes in the structure of the phytoplankton community data and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO)  and changes in the phytoplankton community.
    Our analysis focuses exclusively on bottom-up drivers of community changes, because these data are available throughout the entire time series and at a monthly resolution. Our study builds on and expands substantially the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017. Understanding the regional dynamics of the Cariaco Basin is crucial for unveiling theecological changes affecting this tropical marine ecosystem and the variables driving these changes.

\section{Materials and Methods}
% Here is text on Materials and Methods.
%
\subsection{Study Site}
    %Talk about the Basin shortly, then about the time series
    The Cariaco Ocean Time Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon. The Cariaco Basin (Figure~\ref{fig:map}) is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north, running almost parallel to the coastline, at a mean depth of around \qty{100}{m}. This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at BATS and HOT sites, resulting in the formation of anoxic conditions below \qty{\sim 200}{m} depth.

>>>>

    % Phytoplankton is important
    Phytoplankton provide the foundation for marine food webs and play a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}. Phytoplankton are a paraphyletic group of photo-autotrophic organisms exhibiting highly variable eco-physiological properties \cite{appeltans_magnitude_2012}. The short generation time and high diversity allow phytoplankton communities to respond relatively quickly to environmental changes, making them suitable indicators of ecosystem state \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
    Coastal margins, where phytoplankton typically thrive, constitute only \qty{\sim 10}{\%} of the global ocean's surface but are responsible for an estimated \qty{40}{\%} of carbon sequestration to the seafloor \cite{yool_examination_2001, mullerkarger_importance_2005}.

    Data from remote sensing suggest a global increase in the frequency and spatial extent of coastal phytoplankton blooms over the last 20 years, but regional effects are highly variable \cite{dai_coastal_2023}.
    Climate change is predicted to have a lasting impact on phytoplankton communities, including shifts in community composition and diversity \cite{acevedo-trejos_glimpse_2014, boyd_biological_2016, henson_future_2021}. 
    Long-term time series of biological observations are crucial for understanding how marine ecosystems respond to short-term processes and long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}. Such time series are rare, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.

    A tropical coastal location that has been studied intensively in the last decades is the Cariaco Basin, where a wide range of physical and biogeochemical parameters have been consistently collected from 1995 to 2017 \cite{muller-karger_scientific_2019}. 
    % Describe CARIACO location
    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within one of the largest anoxic basins in the world oceans \cite{edgcomb_accessing_2011}. The basin is a continental shelf depression comprising two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}. The water below the sill is poorly ventilated. This underwater topography in combination with a high surface productivity and aerobic degradation of sinking organic matter, creates permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}. Due to these unique properties, the Cariaco Basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of past climate \cite{hughen1996nature} and serving as a testbed for understanding the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.

    % Describe general seasonal cycle with functional groups etc.
    The Cariaco Basin is characterized by a marked seasonal cycle of upwelling, which is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward. A smaller wind-driven secondary upwelling event occurs often in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}. During the upwelling period, the supply of deeper nutrient-rich waters supports high biological productivity, and the phytoplankton community is dominated by diatoms \cite{romero_seasonal_2009}, with dinoflagellates, coccolithophorids, and nanoplankton also contributing. In the low-wind, rainy season, the Cariaco Basin exhibits oligotrophic conditions, and the fraction of large phytoplankton cells is reduced. However, nano- and picophytoplankton cells are present in relatively high numbers throughout the year \cite{lorenzoni_characterization_2015}.

    % PARAGRAPH THREE - CHANGES IN CARIACO
    Since the inception of the time series in 1995, the ecosystem of the Cariaco Basin has undergone marked changes. \citeA{taylor_ecosystem_2012} analyzed the time series data up to 2010 and found significant decadal trends of increasing sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors attributed to global warming. During the initial phase of the time series (1996–2004), the Cariaco Basin underwent consistent and pronounced seasonal upwelling cycles, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}. After 2004, a significant reduction in the strength of trade winds during the upwelling period \cite{taylor_ecosystem_2012} caused a regime change in the physical and biogeochemical system. This strongly affected the phytoplankton community and the entire ecosystem. Microscopy cell counts indicated a drastic reduction of large phytoplankton, particularly diatoms, as confirmed, albeit not with the same high significance, by analysis of pigment data \cite{pinckney_phytoplankton_2015}. In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, groups commonly found in stratified water columns \cite{lorenzoni_characterization_2015}. An analysis of pigment data indicates a deepening of the chlorophyll maximum, an increase in diversity, and a reduction in the seasonal variability of community composition \cite{pinckney_phytoplankton_2015}.

    % PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
    The shift in phytoplankton community composition associated with increased temperature and reduced upwelling intensities does not appear to be a permanent change in regime. Data from 2012 to 2017, the end phase of the CARIACO time series, indicates a return to strong upwelling intensities and to high numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}.
    The impact of large-scale climatic processes on the observed shifts in the phytoplankton community remains unclear. The Atlantic Multidecadal Oscillation (AMO) has emerged as an important factor influencing the climate system in the region and the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}. The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
    The El Niño Southern Oscillation (ENSO) is teleconnected to the Caribbean Sea, given that during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures increase and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. However, using the time series available only until 2010, \citeA{taylor_ecosystem_2012} found no significant correlation between MEI v.2 and in situ observations, and did not test for correlations with the AMO index.

    % What drives the changes??? Bottom-up/ Top-down ... CLIMATE 
    \citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation".
    Previous analyses of phytoplankton community diversity in the Cariaco Basin did not include time series data beyond 2011, and no detailed investigations have been undertaken yet on possible changes in diversity based on microscopy cell counts.
    Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period from 1995 - 2017, 2) changes in the structure of the phytoplankton community data and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO) and changes in the phytoplankton community.
    Our analysis focuses exclusively on bottom-up drivers of community changes, because these data are available throughout the entire time series and at a monthly resolution. Our study builds on and expands substantially the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017. Understanding the regional dynamics of the Cariaco Basin is crucial for unveiling the ecological changes affecting this tropical marine ecosystem and the variables driving these changes.

\section{Materials and Methods}
% Here is text on Materials and Methods.
%
\subsection{Study Site}
    %Talk about the Basin shortly, then about the time series
    The Cariaco Ocean Time Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon. The Cariaco Basin (Figure~\ref{fig:map}) is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north, running almost parallel to the coastline, at a mean depth of around \qty{100}{m}. This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at BATS and HOT sites, resulting in the formation of anoxic conditions below \qty{\sim 275}{m} depth.

### 2025-06-29 text-backup.tex
Ago Review Intro revised

>>>>

% TOP DOWN ASPECTS FROM INTRO:

During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, coinciding with an increase in zooplankton biomass \cite{taylor_ecosystem_2012}. Interestingly, changes in the phytoplankton community did not result in a significant change in the export of particulate organic carbon (POC) to depth, as measured by sediment traps deployed during the time series \cite{pinckney_phytoplankton_2015}.

The return to conditions similar to those of the late 1990s coincides with a recovery in sardine catches observed in 2016 and 2017 in an area southeast of Isla de Margarita \cite{gomez_gaspar_variacion_2025}.
